---
title: "BellabeatAnalysis"
author: "Shahenaz H"
date: "2024-01-01"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

## Setting up environment
Notes: Setting up environment by loading the packages
```{r loading the packages}
install.packages('tidyverse')
install.packages('ggplot2')
install.packages('here')
install.packages('janitor')
install.packages('dplyr')
install.packages('rmarkdown')
library(tidyverse)
library(ggplot2)
library(here)
library(janitor)
library(dplyr)
library('rmarkdown')
```
## Load CSV files
```{r read all the csv file and create the dataframe}
daily_activity <- read.csv("dailyActivity_merged.csv")
sleep_day <- read.csv("sleepDay_merged.csv")
weight_log_info <- read.csv("weightLogInfo_merged.csv")
daily_steps <- read.csv("dailySteps_merged.csv")
hourly_steps <- read.csv("hourlySteps_merged.csv")
daily_calories <- read.csv("dailyCalories_merged.csv")
```

## Explore a few key tables
```{r Take a look at the dataframe daily_activity.Identify all the columns in the data frame.}
head(daily_activity)
colnames(daily_activity)
```
```{r Take a look at the dataframe sleep_day.Identify all the columns in the data frame.}
head(sleep_day)
colnames(sleep_day)
```
```{r Take a look at the dataframe weight_log_info.Identify all the columns in the data frame.}
head(weight_log_info)

```
```{r Take a look at the dataframe daily_calories.Identify all the columns in the data frame.}
head(daily_calories)
```

```{r Take a look at the dataframe daily_steps.Identify all the columns in the data frame.}
head(daily_steps)
```
```{r Take a look at the dataframe hourly_steps.Identify all the columns in the data frame.}
head(hourly_steps)
```
## Cleaning the data frames
### 1.Removing LoggedActivitiesDistance and SedentaryAvtiveDistance as they are not used in analysis from daily_activity dataframe
```{r Removing LoggedActivitiesDistance and SedentaryAvtiveDistance as they are not used in analysis from daily_activity dataframe}
daily_activity <- subset(daily_activity,select = -c(LoggedActivitiesDistance, SedentaryActiveDistance))
```
```{r Rename column ActivityDate to Date}
colnames(daily_activity)[2] = 'Date'
```
```{r Checking modified daily_activity dataframe}
colnames(daily_activity)
```
### 2. Separating ActivityHour column from hourly_steps dataframe into Date and Hour column for analysis and check modified hourly_steps
```{r Separating ActivityHour column from hourly_steps dataframe into Date and Hour column for analysis and check modified hourly_steps}
hourly_steps <- hourly_steps %>% 
  separate(ActivityHour, c("Date", "Hour"), sep = "^\\S*\\K")
head(hourly_steps)
```
### 3.Removing SleepRecords and time as it is not used in analysis from sleep_day dataframe and check modified dataframe
```{r Removing SleepRecords and time as it is not used in analysis from sleep_day dataframe and check modified dataframe}
sleep_day <- subset(sleep_day,select = -c(TotalSleepRecords))
sleep_day <- sleep_day %>% 
  separate(SleepDay, c("Date"), sep = "^\\S*\\K")
head(sleep_day)
```
## Understanding some summary statistics
### How many unique participants are there in each dataframe? 

```{r Unique participants in daily_activity}
n_distinct(daily_activity$Id)
```
```{r Unique participants in sleep_day}
n_distinct(sleep_day$Id)
```

```{r Unique participants in daily_steps}
n_distinct(daily_steps$Id)
```

```{r Unique participants in daily_calories}
n_distinct(daily_calories$Id)
```

```{r Unique participants in weight_log_info}
n_distinct(weight_log_info$Id)
```
Note: There are only 8 participants in weight_log_info dataframe which is very less to draw any calculation. So we will exclude it from our analysis.

## How many observations are there in each dataframe?
```{r Number of observations in each dataframe}
nrow(daily_activity)
nrow(sleep_day)
nrow(daily_calories)
nrow(daily_steps)
nrow(hourly_steps)

```
##Quick Summary

```{r  For the daily activity dataframe:}
daily_activity %>%  
  select(TotalSteps,
         TotalDistance,
         SedentaryMinutes) %>%
  summary()

```
```{r For the sleep dataframe:}
sleep_day %>%  
  select(TotalMinutesAsleep,
         TotalTimeInBed) %>%
  summary()
```
```{r For the daily_calories dataframe:}
daily_calories %>%
  select(Id,
         Calories) %>%
  summary()
```
```{r For the daily_steps dataframe:}
daily_steps %>%
  select(Id,
         StepTotal) %>%
  summary()
```
```{r For the hourly_steps dataframe}
hourly_steps %>%
  select(Id,
         StepTotal) %>%
  summary()
```
###Finding average steps taken by each participant
```{r from daily_activity}
mean_steps <- daily_activity %>%
  group_by(Id)%>%
  summarise(mean_steps = mean(TotalSteps)) %>%
  select(Id, mean_steps) %>%
  arrange(mean_steps) %>%
  as.data.frame()
head(mean_steps)
```
###Finding average sedentary minutes of each participant
```{r From daily_activity}
mean_sedentary_minutes <- daily_activity %>%
  group_by(Id)%>%
  summarise(mean_sedentary_minutes = mean(SedentaryMinutes)) %>%
  select(Id, mean_sedentary_minutes) %>%
  arrange(mean_sedentary_minutes) %>%
  as.data.frame()
head(mean_sedentary_minutes)
```
###Finding average sleep taken by each participant
```{r From sleep_day}
mean_sleep <- sleep_day %>%
  group_by(Id)%>%
  summarise(mean_sleep = mean(TotalMinutesAsleep)) %>%
  select(Id, mean_sleep) %>%
  arrange(mean_sleep) %>%
  as.data.frame()
head(mean_sleep)
```
###Finding actual minutes of sleep in percentage by each participant
```{r}
actual_sleep <- sleep_day %>%
  group_by(Id) %>%
  mutate(percent_sleep = (TotalMinutesAsleep/TotalTimeInBed)*100) %>%
  select(Id, percent_sleep) %>%
  summarize(actual_sleep = mean(percent_sleep)) %>%
  arrange(actual_sleep) %>%
  mutate_if(is.numeric, round, 2)
  glimpse(actual_sleep)

```
Note : All participants sleep almost 90% of time except four. The least recorded is 63%

### Merging these two datasets together to find additional insights

```{r Combining sleep_day and daily_activity datasets}
combined_data <- merge(sleep_day, daily_activity)
glimpse(combined_data)
```

```{r Combining daily_steps and daily_calories}
combined_steps_calories <- merge(daily_steps, daily_calories)
head(combined_steps_calories)
```
## Data Visualisation
###Plotting a few explorations

```{r }
ggplot(data=daily_activity, aes(x=TotalSteps, y=SedentaryMinutes)) + geom_point() +
  labs(title = "Relation between TotalSteps and SedentaryMinutes ",
       x = "TotalSteps", y = "SedentaryMinutes")
```
Note: The relation between SedentaryMinutes and TotalSteps taken by participant is non-linear. That means TotalSteps taken in a day doesnot depend on sedentary minutes. 

```{r}
ggplot(data=sleep_day, aes(x=TotalMinutesAsleep, y=TotalTimeInBed)) + geom_point() + labs(title = "Relation between TotalMinutesAsleep and TotalTimeInBed ",
       x = "TotalMinutesAsleep", y = "TotalTimeInBed")
```

Note: The relation between TotalMinutesAsleep and TotalTimeInBed is linear. 

```{r}
ggplot(data=combined_steps_calories, aes(x=StepTotal, y=Calories)) + geom_point() + geom_smooth() + labs(title = "Relation between TotalSteps and Calories ",
       x = "StepTotal", y = "Calories")
```

Note: The relation between StepTotal and Calories is increasing gradually. That means the more steps you take, the more calories you burn.

```{r}
ggplot(data=hourly_steps, mapping = aes(x=Hour, y=StepTotal))  +geom_col(fill = 'orange') + theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Average Steps Taken per Hour of Day",
       x = "Hour", y = "Average Steps")
```

Note: Evening 6 P.M, 7 P.M, 10 P.M steps taken are more. whereas morning 10 A.M and afternoon 12 P.M, 2 P.M steps taken are more.


